---
title: "Who Bears the Brunt? Visualizing the Areas Most Impacted by Trump's Grant Cuts"
subtitle: "Project"
author: 
  - name: "Nathaniel Cross"
    affiliations:
      - name: "School of Government and Public Policy, University of Arizona"
description: "This project examines trends in National Science Foundation research grants terminated under the Trump Administration in 2025 adopting a data visualization approach through the creation and interpretation of unique plots."
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: false
  output: false
---

```{r}
#| label: R setup

# Install and load packages
if(!require(pacman))
  install.packages("pacman")

pacman::p_load(tidyverse,
               janitor,
               lubridate,
               scales,
               sysfonts,
               showtext,
               see,
               sf,
               tigris)

devtools::install_github("tidyverse/dsbox")

# Import and load fonts
sysfonts::font_add_google(name = "Source Sans 3", family = "Source Sans 3")
showtext::showtext_auto()

# Set width of code output
options(width = 65)
```

```{r}
#| label: General data loading and cleaning

# Read in dataset
raw_terminations <- read_csv("data/nsf_terminations_airtable.csv")

# Clean data (code sourced from TidyTuesday)
terminations <- raw_terminations |> 
  janitor::clean_names() |> 
  mutate(usaspending_obligated = stringi::stri_replace_first_fixed(usaspending_obligated, "$", "") |> 
           readr::parse_number()) |> 
  mutate(grant_number = as.character(grant_number))

# Date analysis
terminations |>
  glimpse()

terminations |>
  ggplot(aes(x = nsf_startdate)) +
  geom_histogram()
```

## Abstract

## General Introduction

## Question 1

### Introduction

### Approach

### Analysis

```{r}
#| label: Data wrangling for question 1

# Subsetting data and filtering out NAs
term2 <- terminations |>
  select(
    termination_letter_date, 
    directorate, 
    usaspending_obligated
    ) |>
  filter(!if_any(everything(), is.na))

# Distinct directorates
term2 |> select(directorate) |> distinct()

# Filtering out unneeded directorates
term2 <- term2 |>
  filter(directorate != "Office of the Director") 

# Cleaning directorate names
term2 <- term2 |>
  mutate(directorate = str_remove_all(term2$directorate, "\""))

# Summarize data
term2_final <- term2 |>
  group_by(directorate, termination_letter_date) |>
  summarize(cancel_grant = sum(usaspending_obligated), .groups = "drop") |>
  arrange(directorate, termination_letter_date) |>
  group_by(directorate) |>
  mutate(cum_cancel_grant = cumsum(cancel_grant),
         cum_cancel_grant_millions = cum_cancel_grant / 1000000
         )

# Converting directorate to factor
term2_final$directorate = as.factor(as.character(term2_final$directorate))

term2_final <- term2_final |>
  mutate(directorate = fct_rev(fct_relevel( 
           directorate,
           "STEM Education",
           "Mathematical and Physical Sciences",
           "Social, Behavioral and Economic Sciences",
           "Technology, Innovation and Partnerships",
           "Geosciences",
           "Computer and Information Science and Engineering",
           "Biological Sciences",
           "Engineering"
           ))
  )
```

```{r}
#| label: Set ggplot settings

# Set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 20, base_family = "Source Sans 3"))
```

```{r}
#| label: Question 1 plot
#| fig-width: 7
#| fig-height: 14
#| execute:
#| echo: true
#| output: true
#| fig-alt: "The figure is a faceted line graph titled, 'STEM education grants hit hardest by Trump terminations' and depicts canceled National Science Foundation grants throughout the spring of 2025 by directorate. Canceled grants are measured through a cumulative sum, and are tracked across a month-long time period from April 18 to May 15. The data shows that most directorates saw a slight increase in value of canceled grants, especially at the end of April, these directorates mostly plateaued between $50 and $100 million USD in canceled grant value. However, for grants allocated to STEM education organizations or projects, value canceled increased drastically, both at the end of April and the beginning of May, resulting in a cumulaitve $900 million USD in grants terminated for this directorate."

# Plot
term2_final |>
  ggplot(aes(
    x = termination_letter_date, 
    y = cum_cancel_grant_millions, 
    color = directorate)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(vars(directorate), nrow = 8, scales = "free") +
  labs(
    y = "Cumulative sum of terminated grants\n(in millions of USD)\n",
    x = NULL,
    title = "Trump grant terminations hit\nSTEM education initiatives hardest",
    subtitle = "National Science Foundation grants canceled in the spring of 2025, by directorate",
  ) +
  scale_y_continuous(
    breaks = c(0, 450, 900), 
    limits = c(0, 900),
    labels = scales::label_dollar()) +
  scale_x_date(
    limits = as.Date(c("2025-04-18", "2025-05-15")),
    breaks = as.Date(c(
      "2025-04-18", 
      "2025-04-25", 
      "2025-05-02", 
      "2025-05-09", 
      "2025-05-15")),
    date_labels = "%B %d") +
  scale_color_okabeito(order = 1:8, reverse = TRUE) +
  theme(
    plot.title.position = "plot",
    plot.margin = margin(t = 1, l = 1.5, b = 1, r = 1, unit = "lines"),
    legend.position = "none",
    plot.title = element_text(
      face = "bold", 
      size = 55, 
      lineheight = 0.5),
    axis.text.y = element_text(color = "#4f4f4f"),
    axis.text.x = element_text(color = "#4f4f4f"),
    axis.title = element_text(
      color = "black", 
      size = 25, 
      lineheight = 0.5),
    plot.subtitle = element_text(color = "#4f4f4f", size = 25)
  )
```

**Alt text:** The figure is a faceted line graph titled, 'STEM education grants hit hardest by Trump terminations' and depicts canceled National Science Foundation grants throughout the spring of 2025 by directorate. Canceled grants are measured through a cumulative sum, and are tracked across a month-long time period from April 18 to May 15. The data shows that most directorates saw a slight increase in value of canceled grants, especially at the end of April, these directorates mostly plateaued between \$50 and \$100 million USD in canceled grant value. However, for grants allocated to STEM education organizations or projects, value canceled increased drastically, both at the end of April and the beginning of May, resulting in a cumulative \$900 million USD in grants terminated for this directorate.

::: callout-note
Alt text also embedded in source code for enhanced screen-reading capability.
:::

### Discussion

## Question 2

### Introduction

### Approach

### Analysis

```{r}
#| label: Question 2 data wrangling

# Collapse total grant terminations by state
term1 <- terminations |>
  select(org_state, usaspending_obligated) |>
  filter(!is.na(usaspending_obligated)) |>
  group_by(org_state) |>
  summarise(total_canceled_april182025 = sum(usaspending_obligated, na.rm = TRUE))
              
# Double-check for missing values
sapply(term1, function(x) sum(is.na(x)))

terminations |> select(termination_letter_date) |> distinct()
```

```{r}
#| label: Reading and cleaning dataset 2

# Note: this dataset contains all NSF grants, and will be used as a per-state baseline to standardize cancelations

# Read in dataset
raw_nsfgrants <- read_csv("data/all_awards.csv")

# Clean names
grants_all <- raw_nsfgrants |>
  janitor::clean_names() |>
  rename(org_state = state_code_group)

# Removing characters from numeric field
grants_all$award_obligation_to_date <- gsub("[$,.]", "", grants_all$award_obligation_to_date)

# Destringing award values
grants_all$award_obligation_to_date <- as.numeric(as.character(grants_all$award_obligation_to_date))

# Identifying mismatched states
grants_unique <- grants_all |> select(org_state) |> distinct()

grants_unique |>
  anti_join(term1, by = "org_state")

# Cleaning dates
grants_all <- grants_all |> 
  select(
    org_state, 
    effective_start_date, 
    effective_end_date, 
    award_obligation_to_date
    ) |>
  mutate(
    effective_start_date = as.Date(effective_start_date, format = "%m/%d/%Y"),
    effective_end_date = as.Date(effective_end_date, format = "%m/%d/%Y")
    ) |>
  filter(effective_start_date >= as.Date("2010-10-01") & 
         effective_start_date <= as.Date("2024-09-30"))

# Extract start year from date
grants_all$start_year <- as.numeric(format(grants_all$effective_start_date,'%Y'))

# Filtering out states not present in the terminations dataset
grants_all <- grants_all |>
  filter(
    org_state != "Non States" &
    org_state != "ON" &
    org_state != "WY" & 
    org_state != "GU" & 
    org_state != "BC"
    )

# Calculating summary statistics
grants_all <- grants_all |>
  group_by(org_state, start_year) |>
  summarize(total_open_perstateyear = sum(
    award_obligation_to_date, 
    na.rm = TRUE)) |>
  group_by(org_state) |>
  summarize(total_open_avgperstate = mean(total_open_perstateyear))
```

```{r}
#| label: Reading in partisanship df

states_vote <- read_csv("data/states_vote.csv") |>
  rename(org_state = state) |>
  mutate(vote_trump = as.logical(vote_trump))
```

```{r}
#| label: Final dataset preparation

# Merging datasets
term1_merged <- term1 |>
  left_join(grants_all, by = "org_state") |>
  left_join(states_vote, by = "org_state") |>
  filter(!is.na(vote_trump))

# Calculating ratio and manipulating based on partisanship
term1_final <- term1_merged |>
  mutate(
    cancel_ratio = total_canceled_april182025 / total_open_avgperstate,
    cancel_ratio = case_when(
      vote_trump == 1 ~ -cancel_ratio,
      TRUE            ~ cancel_ratio
    ),
    cancel_ratio_perc = cancel_ratio * 100
  )

# Set tigris options
options(tigris_class = "sf")

# Loading in data
states <- states(cb = TRUE, resolution = "20m") |>
  filter(NAME != "Puerto Rico")

# Shifting AK and HI closer to the continent and resizing
states_shifted <- shift_geometry(states)

# Merging grant terminations onto base map
term_map <- states_shifted |>
  left_join(term1_final, by = c("STUSPS" = "org_state"), multiple = "all") |>
  replace_na(list(cancel_ratio_perc = 0))
# Note: only na was Wyoming, and was due to the fact that no grants in this state were canceled.
```

```{r}
#| label: Reset parameters and theme for plot

# Set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_void(base_size = 20, base_family = "Source Sans 3"))
```

```{r}
#| label: Question 2 map plot
#| fig-width: 7
#| fig-asp: 0.618
#| execute:
#| echo: true
#| output: true
#| fig-alt: "The figure is a chloropleth titled, 'Massachusetts, Arkansas, and Alabama lost greatest proportions of grant funding in Trump terminations' and depicts the severity of National Science Foundation grant terminations by state and state partisanship. All fifty U.S. states are pictured, and states' fill is mapped to both partisanship and grant terminations. Terminations are calculated as a proportion of value of grants terminated to average annual allocation to each state. Arkansas and Alabama are the two Republican states to suffer the worst from terminations, with canceled value around 0.6% of average annual allocations. Contrastingly, Massachusetts is the Democratic state most affected by the cancellations, with a ratio of around 0.8%."

# Plot
term_map |>
  ggplot()+
  geom_sf(aes(fill = cancel_ratio_perc), color = "black") +
  coord_sf(clip = "off") +
  labs(
    title = "Massachusetts, Arkansas, and Alabama\nlost greatest proportions of grant funding\nin Trump terminations",
    subtitle = "Severity of  National Science Foundation grant terminations, by partisanship",
    fill = "Proportion of terminated\ngrants relative to\naverage annual allocation\n(diverging on state partisanship)",
      ) +
  scale_fill_gradient2(
    low = "#BB1C1C", 
    mid = "white", 
    high = "#0072B2", 
  labels = function(x) 
    label_number(suffix = "%", accuracy = 0.1)
    (abs(x)),
    limits = c(-0.8, 0.8)
    ) +
  guides(fill = guide_colorbar(ticks.colour = NA)) +
  theme(plot.title.position = "plot",
        plot.title = element_text(
         face = "bold", 
         size = 45, 
         lineheight = 0.5),
        legend.title = element_text(
          lineheight = 0.5, 
          color = "#4f4f4f"),
        legend.text = element_text(color = "#4f4f4f"),
        plot.subtitle = element_text(color = "#4f4f4f", size = 25),
       )
```

**Alt text:** The figure is a chloropleth titled, 'Massachusetts, Arkansas, and Alabama lost greatest proportions of grant funding in Trump terminations' and depicts the severity of National Science Foundation grant terminations by state and state partisanship. All fifty U.S. states are pictured, and states' fill is mapped to both partisanship and grant terminations. Terminations are calculated as a proportion of value of grants terminated to average annual allocation to each state. Arkansas and Alabama are the two Republican states to suffer the worst from terminations, with canceled value around 0.6% of average annual allocations. Contrastingly, Massachusetts is the Democratic state most affected by the cancellations, with a ratio of around 0.8%.

::: callout-note
Alt text also embedded in source code for enhanced screen-reading capability.
:::

### Discussion

## Conclusion
