---
title: "Who Bears the Brunt?"
subtitle: "Visualizing the Areas Most Impacted by Trump's Grant Cuts"
author: "Nathaniel Cross"
title-slide-attributes:
  data-background-size: stretch
  data-background-opacity: "0.7"
  data-slide-number: none
format:
  revealjs:
    theme: dark
    smaller: true
    scrollable: true
    incremental: true
editor: visual
execute:
  echo: false
---

```{r}
#| label: Load packages
#| include: false

#     theme:  ['data/customtheming.scss']


# Install and load packages
if(!require(pacman))
  install.packages("pacman")

pacman::p_load(tidyverse,
               janitor,
               lubridate,
               scales,
               sysfonts,
               showtext,
               see,
               sf,
               tigris)

devtools::install_github("tidyverse/dsbox")

# Import and load fonts
sysfonts::font_add_google(name = "Source Sans 3", family = "Source Sans 3")
showtext::showtext_auto()
```

```{r}
#| label: Setting up ggplot
#| include: false

# Plot theme
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11, base_family = "Source Sans 3"))
```

```{r}
#| label: General data loading and cleaning
#| include: false

# Read in dataset
raw_terminations <- read_csv("data/nsf_terminations_airtable.csv")

# Clean data (code sourced from TidyTuesday)
terminations <- raw_terminations |> 
  janitor::clean_names() |> 
  mutate(usaspending_obligated = stringi::stri_replace_first_fixed(usaspending_obligated, "$", "") |> 
           readr::parse_number()) |> 
  mutate(grant_number = as.character(grant_number)) |>
  select(
    termination_letter_date, 
    directorate, 
    usaspending_obligated,
    org_state, 
    nsf_startdate
    )
```

```{r}
#| label: Data wrangling for question 1
#| include: false

# Subsetting data and filtering out NAs
term2 <- terminations |>
  select(
    termination_letter_date, 
    directorate, 
    usaspending_obligated
    ) |>
  filter(!if_any(everything(), is.na))

# Distinct directorates
term2 |> select(directorate) |> distinct()

# Filtering out unneeded directorates
term2 <- term2 |>
  filter(directorate != "Office of the Director") 

# Cleaning directorate names
term2 <- term2 |>
  mutate(directorate = str_remove_all(term2$directorate, "\""))

# Summarize data
term2_final <- term2 |>
  group_by(directorate, termination_letter_date) |>
  summarize(cancel_grant = sum(usaspending_obligated), .groups = "drop") |>
  arrange(directorate, termination_letter_date) |>
  group_by(directorate) |>
  mutate(cum_cancel_grant = cumsum(cancel_grant),
         cum_cancel_grant_millions = cum_cancel_grant / 1000000
         )

# Converting directorate to factor
term2_final$directorate = as.factor(as.character(term2_final$directorate))

term2_final <- term2_final |>
  mutate(directorate = fct_rev(fct_relevel( 
           directorate,
           "STEM Education",
           "Mathematical and Physical Sciences",
           "Social, Behavioral and Economic Sciences",
           "Technology, Innovation and Partnerships",
           "Geosciences",
           "Computer and Information Science and Engineering",
           "Biological Sciences",
           "Engineering"
           ))
  )
```

```{r}
#| label: Question 1 data wrangling for raw count
#| include: false

# Summarize data
term2_final_count <- term2 |>
  group_by(directorate, termination_letter_date) |>
  summarize(cancel_grant_quant = n(), .groups = "drop") |>
  arrange(directorate, termination_letter_date) |>
  group_by(directorate) |>
  mutate(cum_cancel_grant_quant = cumsum(cancel_grant_quant))

# Converting directorate to factor
term2_final_count$directorate = as.factor(as.character(term2_final_count$directorate))

term2_final_count <- term2_final_count |>
  mutate(directorate = fct_rev(fct_relevel( 
           directorate,
           "STEM Education",
           "Mathematical and Physical Sciences",
           "Social, Behavioral and Economic Sciences",
           "Technology, Innovation and Partnerships",
           "Geosciences",
           "Computer and Information Science and Engineering",
           "Biological Sciences",
           "Engineering"
           ))
  )
```

```{r}
#| label: Question 2 data wrangling
#| include: false

# Collapse total grant terminations by state
term1 <- terminations |>
  select(org_state, usaspending_obligated) |>
  filter(!is.na(usaspending_obligated)) |>
  group_by(org_state) |>
  summarise(total_canceled_april182025 = sum(usaspending_obligated, na.rm = TRUE))
```

```{r}
#| label: Reading and cleaning dataset 2
#| include: false

# Note: this dataset contains all NSF grants, and will be used as a per-state baseline to standardize cancelations

# Read in dataset
raw_nsfgrants <- read_csv("data/all_awards.csv")

# Clean names
grants_all <- raw_nsfgrants |>
  janitor::clean_names() |>
  rename(org_state = state_code_group)

# Removing characters from numeric field
grants_all$award_obligation_to_date <- gsub("[$,.]", "", grants_all$award_obligation_to_date)

# Destringing award values
grants_all$award_obligation_to_date <- as.numeric(as.character(grants_all$award_obligation_to_date))

# Identifying mismatched states
grants_unique <- grants_all |> select(org_state) |> distinct()

grants_unique |>
  anti_join(term1, by = "org_state")

# Cleaning dates
grants_all <- grants_all |> 
  select(
    org_state, 
    effective_start_date, 
    effective_end_date, 
    award_obligation_to_date
    ) |>
  mutate(
    effective_start_date = as.Date(effective_start_date, format = "%m/%d/%Y"),
    effective_end_date = as.Date(effective_end_date, format = "%m/%d/%Y")
    ) |>
  filter(effective_start_date >= as.Date("2010-10-01") & 
         effective_start_date <= as.Date("2024-09-30"))

# Extract start year from date
grants_all$start_year <- as.numeric(format(grants_all$effective_start_date,'%Y'))

# Filtering out states not present in the terminations dataset
grants_all <- grants_all |>
  filter(
    org_state != "Non States" &
    org_state != "ON" &
    org_state != "WY" & 
    org_state != "GU" & 
    org_state != "BC"
    )

# Calculating summary statistics
grants_all <- grants_all |>
  group_by(org_state, start_year) |>
  summarize(total_open_perstateyear = sum(
    award_obligation_to_date, 
    na.rm = TRUE)) |>
  group_by(org_state) |>
  summarize(total_open_avgperstate = mean(total_open_perstateyear))
```

```{r}
#| label: Reading in partisanship df
#| include: false

states_vote <- read_csv("data/states_vote.csv") |>
  rename(org_state = state) |>
  mutate(vote_trump = as.logical(vote_trump))
```

```{r}
#| label: Final dataset preparation
#| include: false

# Merging datasets
term1_merged <- term1 |>
  left_join(grants_all, by = "org_state") |>
  left_join(states_vote, by = "org_state") |>
  filter(!is.na(vote_trump))

# Calculating ratio and manipulating based on partisanship
term1_final <- term1_merged |>
  mutate(
    cancel_ratio = total_canceled_april182025 / total_open_avgperstate,
    cancel_ratio = case_when(
      vote_trump == 1 ~ -cancel_ratio,
      TRUE            ~ cancel_ratio
    ),
    cancel_ratio_perc = cancel_ratio * 100
  )

# Set tigris options
options(tigris_class = "sf")

# Loading in data
states <- states(cb = TRUE, resolution = "20m") |>
  filter(NAME != "Puerto Rico")

# Shifting AK and HI closer to the continent and resizing
states_shifted <- shift_geometry(states)
states_shifted <- st_transform(states_shifted, crs = 4326)

# Merging grant terminations onto base map
term_map <- states_shifted |>
  left_join(term1_final, by = c("STUSPS" = "org_state"), multiple = "all") |>
  replace_na(list(cancel_ratio_perc = 0))
# Note: only na was Wyoming, and was due to the fact that no grants in this state were canceled.
```

## Introduction

-   Examination of grants terminated by the Trump Administration

-   Visualizations help understand who is most impacted by these terminations

    -   Scientific domains

    -   Geographic entities

-   Tracks grants housed under the National Science Foundation

## National Science Foundation Grant Terminations

-   Derived from Grant Watch

    -   Independent project tracking the termination of scientific research grants under the Trump Administration's second term

    -   Data is sourced largely by submissions from affected investigators

    -   Tracks individual grant terminations, and includes information on the organization receiving the grant, date of termination, and value terminated

    -   Time frame spans from April 18 to May 15

-   Supplementary data from the National Science Foundation and CNN Politics

## Question 1

**Which scientific domains saw the highest grant termination rates, and how do these terminations vary over time?**

::::: columns
::: {.column width="50%"}
Introduction

-   With a wide impacts of grant terminations, I wanted to explore which scientific domains have been the most impacted

-   Local impacts here at the University of Arizona

-   NSF is divided into directorates, administrative units that manage grants in certain fields
:::

::: {.column width="50%"}
Approach

-   Faceted line plots to best represent time-series data and differences between directorates

-   Use of cumulative summations to depict total grant terminations

-   For reference, both dollar value and quantity are plotted in the project
:::
:::::

------------------------------------------------------------------------

::: panel-tabset
## Introductory subset

```{r}
#| label: Question 1 plot code subsetted
#| execute:
#| echo: false

# Plot
term2_final |>
  filter(directorate == "Engineering") |>
  ggplot(aes(
    x = termination_letter_date, 
    y = cum_cancel_grant_millions, 
    color = directorate)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(vars(directorate), nrow = 4, scales = "free") +
  labs(
    y = "Cumulative sum of terminated grants\n(in millions of USD)\n",
    x = NULL,
    title = "Trump grant terminations hit\nSTEM education initiatives hardest",
    subtitle = "National Science Foundation grants canceled in the spring of 2025, by directorate",
  ) +
  scale_y_continuous(
    breaks = c(0, 450, 900), 
    limits = c(0, 900),
    labels = scales::label_dollar()) +
  scale_x_date(
    limits = as.Date(c("2025-04-18", "2025-05-15")),
    breaks = as.Date(c(
      "2025-04-18", 
      "2025-04-25", 
      "2025-05-02", 
      "2025-05-09", 
      "2025-05-15")),
    date_labels = "%B %d") +
  scale_color_okabeito(order = 1:8, reverse = TRUE) +
  theme(
    plot.title.position = "plot",
    plot.margin = margin(t = 1, l = 1.5, b = 1, r = 1, unit = "lines"),
    legend.position = "none",
    plot.title = element_text(
      face = "bold", 
      size = 55, 
      lineheight = 0.5),
    strip.text = element_text(size = 20),
    axis.text.y = element_text(color = "#4f4f4f", size = 20),
    axis.text.x = element_text(color = "#4f4f4f", size = 20),
    axis.title = element_text(
      color = "black", 
      size = 25, 
      lineheight = 0.5),
    plot.subtitle = element_text(color = "#4f4f4f", size = 25)
  )
```

## Plot

```{r}
#| label: Question 1 plot code
#| execute:
#| echo: false

# Plot
term2_final |>
  ggplot(aes(
    x = termination_letter_date, 
    y = cum_cancel_grant_millions, 
    color = directorate)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(vars(directorate), nrow = 4, scales = "free") +
  labs(
    y = "Cumulative sum of terminated grants\n(in millions of USD)\n",
    x = NULL,
    title = "Trump grant terminations hit\nSTEM education initiatives hardest",
    subtitle = "National Science Foundation grants canceled in the spring of 2025, by directorate",
  ) +
  scale_y_continuous(
    breaks = c(0, 900), 
    limits = c(0, 900),
    labels = scales::label_dollar()) +
  scale_x_date(
    limits = as.Date(c("2025-04-18", "2025-05-15")),
    breaks = as.Date(c(
      "2025-04-18", 
      "2025-04-25", 
      "2025-05-02", 
      "2025-05-09", 
      "2025-05-15")),
    date_labels = "%B %d") +
  scale_color_okabeito(order = 1:8, reverse = TRUE) +
  theme(
    plot.title.position = "plot",
    plot.margin = margin(t = 1, l = 1.5, b = 1, r = 1, unit = "lines"),
    legend.position = "none",
    plot.title = element_text(
      face = "bold", 
      size = 55, 
      lineheight = 0.5),
    strip.text = element_text(size = 20),
    axis.text.y = element_text(color = "#4f4f4f", size = 20),
    axis.text.x = element_text(color = "#4f4f4f", size = 20),
    axis.title = element_text(
      color = "black", 
      size = 25, 
      lineheight = 0.5),
    plot.subtitle = element_text(color = "#4f4f4f", size = 25)
  )
```

## STEM Education subset

```{r}
#| label: Question 1 plot code STEM subset
#| execute:
#| echo: false

# Plot
term2_final |>
  filter(directorate == "STEM Education") |>
  ggplot(aes(
    x = termination_letter_date, 
    y = cum_cancel_grant_millions, 
    color = directorate)) +
  geom_line(linewidth = 1.1, color = "#E69F00") +
  geom_point(size = 2, color = "#E69F00") +
  facet_wrap(vars(directorate), nrow = 4, scales = "free") +
  labs(
    y = "Cumulative sum of terminated grants\n(in millions of USD)\n",
    x = NULL,
    title = "Trump grant terminations hit\nSTEM education initiatives hardest",
    subtitle = "National Science Foundation grants canceled in the spring of 2025, by directorate",
  ) +
  scale_y_continuous(
    breaks = c(0, 450, 900), 
    limits = c(0, 900),
    labels = scales::label_dollar()) +
  scale_x_date(
    limits = as.Date(c("2025-04-18", "2025-05-15")),
    breaks = as.Date(c(
      "2025-04-18", 
      "2025-04-25", 
      "2025-05-02", 
      "2025-05-09", 
      "2025-05-15")),
    date_labels = "%B %d") +
  theme(
    plot.title.position = "plot",
    plot.margin = margin(t = 1, l = 1.5, b = 1, r = 1, unit = "lines"),
    legend.position = "none",
    plot.title = element_text(
      face = "bold", 
      size = 55, 
      lineheight = 0.5),
    strip.text = element_text(size = 20),
    axis.text.y = element_text(color = "#4f4f4f", size = 20),
    axis.text.x = element_text(color = "#4f4f4f", size = 20),
    axis.title = element_text(
      color = "black", 
      size = 25, 
      lineheight = 0.5),
    plot.subtitle = element_text(color = "#4f4f4f", size = 25)
  )
```
:::

## Question 2

**Were grant terminations clustered by geography or motivated by states' political leanings?**

::::: columns
::: {.column width="50%"}
Introduction

-   Motivated by variation in state and federal policy, especially in how states implement policy

-   Era of record high interstate and partisan polarization

-   Map plots are cool :)
:::

::: {.column width="50%"}
Approach

-   Color-mapped choropleth

    -   Hue mapped to states' partisanship

    -   Value mapped to grant terminations

-   Grant terminations measured in dollar value as a proportion of average annual grant allocation through the NSF to standardize differences between states

-   Partisanship measured through proxy of 2024 presidential election results
:::
:::::

------------------------------------------------------------------------

::: panel-tabset
## Plot

```{r}
#| label: Question 2 map plot all US
#| execute:
#| echo: false

# Plot
term_map |>
  ggplot()+
  geom_sf(aes(fill = cancel_ratio_perc), color = "black") +
  coord_sf(clip = "off") +
  labs(
    title = "Massachusetts and D.C.lost greatest proportions\nof grant funding in Trump terminations",
    subtitle = "Severity of  National Science Foundation grant terminations, by partisanship",
    fill = "Proportion of terminated\ngrants relative to\naverage annual allocation\n(diverging on state partisanship)\n",
      ) +
  scale_fill_gradient2(
    low = "#BB1C1C", 
    mid = "white", 
    high = "#0072B2", 
  labels = function(x) 
    label_number(suffix = "%", accuracy = 0.1)
    (abs(x)),
    limits = c(-0.8, 0.8)
    ) +
  guides(fill = guide_colorbar(ticks.colour = NA)) +
  theme(plot.title.position = "plot",
        plot.title = element_text(
         face = "bold", 
         size = 45, 
         lineheight = 0.5),
        legend.title = element_text(
          lineheight = 0.5, 
          color = "#4f4f4f",
          size = 20),
        legend.text = element_text(color = "#4f4f4f", size = 20),
        plot.subtitle = element_text(color = "#4f4f4f", size = 25),
       )
```

## Northeast detail

```{r}
#| label: Q2 plot NE
#| execute:
#| echo: false

# Plot
term_map |>
  ggplot()+
  geom_sf(aes(fill = cancel_ratio_perc), color = "black") +
  coord_sf(xlim = c(-79, -69), ylim = c(37, 44), expand = TRUE) +
  labs(
    title = "Massachusetts and D.C.lost greatest proportions\nof grant funding in Trump terminations",
    subtitle = "Severity of  National Science Foundation grant terminations, by partisanship",
    fill = "Proportion of terminated\ngrants relative to\naverage annual allocation\n(diverging on state partisanship)\n",
      ) +
  scale_fill_gradient2(
    low = "#BB1C1C", 
    mid = "white", 
    high = "#0072B2", 
  labels = function(x) 
    label_number(suffix = "%", accuracy = 0.1)
    (abs(x)),
    limits = c(-0.8, 0.8)
    ) +
  guides(fill = guide_colorbar(ticks.colour = NA)) +
  theme(plot.title.position = "plot",
        plot.title = element_text(
         face = "bold", 
         size = 45, 
         lineheight = 0.5),
        legend.title = element_text(
          lineheight = 0.5, 
          color = "#4f4f4f",
          size = 20),
        legend.text = element_text(color = "#4f4f4f", size = 20),
        plot.subtitle = element_text(color = "#4f4f4f", size = 25),
       )
```
:::

## Conclusion

::::: columns
::: {.column width="50%"}
-   Many losers, no winners

-   Universities and research institutions everywhere impacted

    -   Especially in states like Massachusetts, D.C., Maryland, Alabama, and Arkansas

    -   And institutions focusing on STEM Education or similar initiatives

-   Long-lasting consequences on research and scientific enquiry
:::

::: {.column width="50%"}
-   Limitations

    -   Presentation of raw or summarized data

    -   Standardizing across states in Question 2

-   Further directions:

    -   Continued analysis as terminations continue

    -   Critical to track and document this data as it becomes available
:::
:::::

# Thank you
